---
title: "Introduction to functional programming in R"
author: "Cecile Murray, Data Scientist, ERD BDS"
date: "10/27/2021"
output:
  beamer_presentation: default
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## Goals

- What is functional programming and why is it useful?
- Describe the components of a function
- Explore how to use functions to reduce repetition via demo

## What is functional programming?

*“When using a functional style, you strive to decompose components of the problem into isolated functions that operate independently. Each function taken by itself is simple and straightforward to understand; complexity is handled by composing functions in various ways.”*

- Hadley Wickham, Advanced R  

## How is functional programming different?

- Procedural programming = repeatedly writing and executing sequential commands 
- Functional programming = writing functions to define a set of operations, then repeatedly executing the functions 

## Functional principles apply outside of R

- R 
- Python
- Stata (macros)
- SAS (macros)
- many, many more

## What is a function?

A function is a canned bit of code that takes in some inputs, performs some operations, and spits out some outputs.

Example: `mean()`

\footnotesize
```{r mean}
digits <- rnorm(10)
digits

mean(digits)
```

## Example: procedural code

\footnotesize
```{r procedural_example}
df <- tibble::tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

# rescale from 0-1
df$a <- (df$a - min(df$a, na.rm = TRUE)) / 
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
df$b <- (df$b - min(df$b, na.rm = TRUE)) / 
  (max(df$b, na.rm = TRUE) - min(df$a, na.rm = TRUE))
df$c <- (df$c - min(df$c, na.rm = TRUE)) / 
  (max(df$c, na.rm = TRUE) - min(df$c, na.rm = TRUE))
df$d <- (df$d - min(df$d, na.rm = TRUE)) / 
  (max(df$d, na.rm = TRUE) - min(df$d, na.rm = TRUE))

```

\footnotesize Source: R for Data Science

## Example: procedural code

\footnotesize
```{r procedural_example_2}
df <- tibble::tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

# rescale from 0-1
df$a <- (df$a - min(df$a, na.rm = TRUE)) / 
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE)) 
df$b <- (df$b - min(df$b, na.rm = TRUE)) / 
  (max(df$b, na.rm = TRUE) - min(df$a, na.rm = TRUE)) # copy-paste bug
df$c <- (df$c - min(df$c, na.rm = TRUE)) / 
  (max(df$c, na.rm = TRUE) - min(df$c, na.rm = TRUE))
df$d <- (df$d - min(df$d, na.rm = TRUE)) / 
  (max(df$d, na.rm = TRUE) - min(df$d, na.rm = TRUE))

```

Source: R for Data Science

## Example: functional code

\footnotesize
```{r functional_example}
# define function that rescales values from 0-1
rescale <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

# call the function on each column
df$a <- rescale(df$a)
df$b <- rescale(df$b)
df$c <- rescale(df$c)
df$d <- rescale(df$d)
```

<font size="1">Source: R for Data Science</font>

## Benefits of a functional approach

Functions can:

- Reduce repetition/duplication in code
- Enable testing to ensure code produces expected output
- Serve as modular building blocks that can be re-used elsewhere  

## DRY programming: "Don't repeat yourself" 

Duplicated code is:  

- Harder for other people to understand  
- More likely to contain copy and paste or other "human" errors  
- Harder to parse for errors in logic or unexpected side effects 
- Harder to maintain over time 
- Possibly slower to run (probably only relevant on very large datasets)  

## Functions can be tested

- Verify that your function produces expected outputs for given inputs
- Know with certainty that you will get correct output anywhere you use it
- Much harder and more labor-intensive to do this with procedural code

## Reusability

- Once you write a function, you can use it again anywhere you want
- This also allows you to use functions as building blocks

## Key elements of functions

- name
- inputs, also known as arguments
- output, often called return value
- body: where computation is defined

## Key elements of functions

```{r func, eval=FALSE}
do_something <- function(input1, input2) {
  out <- input1 %>% some_operation(input2) 
  return(out)
}

```

## How do we avoid repetitive code?

\footnotesize
```{r repeated}
# call the function on each column
df$a <- rescale(df$a)
df$b <- rescale(df$b)
df$c <- rescale(df$c)
df$d <- rescale(df$d)
```

Options:

  - loops
  - base R `apply()` functions
  - `purrr::map()`

## Demo

## Some topics we didn't cover

Function topics:

- scope + return values vs. side effects
- lazy evaluation: arguments don't get evaluated until they're called
- arbitrary number of inputs using `...`


```purrr```

- `purrr::map2` for two inputs
- `purrr::pmap` for arbitrary number of inputs
- list-columns: a column where each cell is a list instead of a single value (good for simulations/modeling)

## Thank you! Questions?

More detail: 

- R4DS Ch. 19 https://r4ds.had.co.nz/functions.html
- R4DS Ch. 21 https://r4ds.had.co.nz/iteration.html
- https://purrr.tidyverse.org/index.html

Contact me: cecile.m.murray@census.gov


